---
title: "FA 5 – Probability and Distributions"
author: "Amyah I. Espanol"
output:
  pdf_document:
    toc: false
    number_sections: true
fontsize: 12pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**https://github.com/amyahespanol/Amyah-Espanol-Statistical-Theory-APM1111.git**

# Problem 1

Draw three cards in succession from a deck of 52 cards.
Let:

- $E_1$ = King on the first draw  
- $E_2$ = King on the second draw  
- $E_3$ = King on the third draw  

## (a) $P(\overline{E_1E_2})$

The expression $\overline{E_1E_2}$ gives the probabilty of the intersection where at least ONE of the first two draws are NOT a king.

## (b) $P(E_1 + E_2)$

The symbol $E_1 + E_2$ is the union $E_1 \cup E_2$.
$P(E_1 + E_2)$ is the probability that a king is drawn on the first draw, OR on the second draw, OR on both.

## (c) $\overline{E_1 + E_2}$

This is the event that neither the first nor the second draw is a king.

## (d) $P(E_3 \mid E_1\ \overline{E_2})$

This represents the conditional probability that a king is drawn on the third draw,
given that the first draw was a king and the second one was NOT.

## (e) $\overline{E_1E_2E_3}$

This gives the intersection $\overline{E_1} \cap \overline{E_2} \cap \overline{E_3}$, so a king is NOT drawn on all three draws.

## (f) $P(E_1E_2 + E_2\ \overline{E_2})$

This represents the probability that, 
A king is drawn on both the first and second draws OR A king is NOT drawn on the second draw and a king is drawn on the third.

# Problem 2

## (a) Probability Distribution of $X$

Three fair dice are rolled. Let $X$ be the sum of the three upper faces.

The dice are independent, giving it 216 total outcomes.
$$
\text{Total number of outcomes} = 6^3 = 216.
$$

To determine the probability distribution of $X$, I used R.
```{r}
# 1. Generate all 216 possible outcomes for three dice
dice_grid <- expand.grid(d1 = 1:6, d2 = 1:6, d3 = 1:6)

# 2. Calculate the sum (X) for each outcome
dice_grid$X <- rowSums(dice_grid)

# 3. Create a frequency table
freq_data <- as.data.frame(table(dice_grid$X))
colnames(freq_data) <- c("x", "Frequency")

# 4. Calculate probabilities
freq_data$Probability <- as.numeric(freq_data$Frequency) / 216

# 5. Markdown table
knitr::kable(freq_data, 
             col.names = c("Sum (x)", "Frequency", "P(X = x)"),
             caption = "Probability Distribution of the Sum of Three Dice",
             digits = 4)
```

Therefore, the probability distribution of $X$ is:
$$
P(X = x) = \frac{\text{Number of outcomes that sum to } x}{216},
$$
where $x = 3,4,\dots,18$.

The distribution has symmetry around 10.5.

## (b) Compute $P(7 \le X \le 11)$

Compute:
$$
P(7 \le X \le 11) = P(X=7) + P(X=8) + P(X=9) + P(X=10) + P(X=11).
$$
```{r}
# Compute probability

prob_7_11 <- sum(freq_data$Probability[as.numeric(as.character(freq_data$x)) >= 7 & as.numeric(as.character(freq_data$x)) <= 11])

prob_7_11
```

Therefore,
$$
P(7 \le X \le 11) = `r prob_7_11`.
$$

# Problem 3

Given distribution find these probabilities.  

X <- c(-10, -20, 30)
p <- c(1/5, 3/10, 1/2)

## (a) E(X)
EX <- sum(X * p) = 7

## (b) E(X^2)
EX2 <- sum((X^2) * p) = 590

## (c) Variance = E(X^2) - (E(X))^2
VarX <- EX2 - EX^2 = 541

## (d) E(X^3)
EX3 <- sum((X^3) * p) = 10900

# Problem 4

In how many ways can 3 men and 3 women be seated at a round table if:

Let the six people be distinct individuals.

```{r problem4, echo=TRUE}

# (a) No restriction: 
# Number of circular permutations of 6 different people = (6-1)!
a <- factorial(6-1)

# (b) Two particular women must not sit together:
# Total arrangements - arrangements where W1 and W2 sit together
total <- a
# keep W1 and W2 as one -> 5 people around a circle -> (5-1)!; multiply by 2
together <- factorial(5-1) * 2
not_together <- total - together

# (c) Each woman is to be between two men (men and women alternate).
# Number of alternating seatings, men first around the circle: (3-1)!
# Place the 3 distinct women in the 3 gaps: 3!
alternating <- factorial(3-1) * factorial(3)

list(
  a_no_restriction = a,
  b_not_together = not_together,
  c_alternating = alternating
)
```

## (a)
Number of arrangements with no restriction: `r a`

## (b)
Number of arrangements where two particular women are not together: `r not_together`

## (c)
Number of alternating seatings (each woman between two men): `r alternating`

# Problem 5

In how many ways can 2 men, 4 women, 3 boys, and 3 girls be selected from:  
6 men, 8 women, 4 boys, 5 girls

```{r problem5, echo=TRUE}
choose <- choose

# (a) No restrictions
a_men <- choose(6,2)
a_women <- choose(8,4)
a_boys <- choose(4,3)
a_girls <- choose(5,3)
total_a <- a_men * a_women * a_boys * a_girls

# (b) A particular man and a particular woman must be selected.

# For men: 2 men needed total, 1 is fixed -> choose 1 from remaining 5
b_men <- choose(5,1)
# For women: 4 women needed total, 1 is fixed -> choose 3 from remaining 7
b_women <- choose(7,3)
b_boys <- a_boys
b_girls <- a_girls
total_b <- b_men * b_women * b_boys * b_girls

list(
  total_a = total_a,
  total_b = total_b
)
```
## (a)
No restriction: `r total_a`

## (b)
A particular man and a particular woman must be selected: `r total_b`

# Problem 6

Prove that for any two events \(E_1\) and \(E_2\),
\[
P(E_1 + E_2) \le P(E_1) + P(E_2)
\]

---

## Solution

Using the Addition Rules for probabilities:
\[
P(E_1 + E_2) = P(E_1) + P(E_2) - P(E_1E_2)
\]

Since the probability of an intersection can't be negative,
\[
P(E_1E_2) \ge 0
\]

Subtracting a nonnegative number can't make the result larger, so
\[
P(E_1 + E_2)
  = P(E_1) + P(E_2) - P(E_1E_2)
  \le P(E_1) + P(E_2)
\]

Therefore,
\[
\boxed{P(E_1 + E_2) \le P(E_1) + P(E_2)}
\]

**Conclusion:**  
The probability of the union of two events is always less than or equal to the sum of their probabilities.